---
title: "Automatic model diagrams"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Automatic model diagrams}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
hasDiagrammeR <- requireNamespace("DiagrammeR", quietly = TRUE)
```

A compartment diagram is often the fastest way to explain a model in a report
or to check that the equations say what you meant.  `nlmixr2plot` can draw one
directly from a model's differential equations:

- `modelGraph()` parses the equations into a graph of compartments and flows.
- `modelDiagram()` lays that graph out following common pharmacometric
  conventions and draws it.

```{r setup}
library(nlmixr2plot)
```

## A first diagram

Any model function, `rxode2` model, `rxode2` user interface object or fitted
`nlmixr2` object can be diagrammed.  Here is a two-compartment model with
first-order absorption:

```{r}
two.cmt <- function() {
  ini({
    tka <- log(1.5)
    tcl <- log(3)
    tv <- log(20)
    tq <- log(2)
    tvp <- log(40)
    add.sd <- 0.2
  })
  model({
    ka <- exp(tka)
    cl <- exp(tcl)
    v <- exp(tv)
    q <- exp(tq)
    vp <- exp(tvp)
    d/dt(depot) <- -ka * depot
    d/dt(central) <- ka * depot - cl / v * central - q / v * central +
      q / vp * periph
    d/dt(periph) <- q / v * central - q / vp * periph
    cp <- central / v
    cp ~ add(add.sd)
  })
}
```

```{r}
modelDiagram(two.cmt, engine = "ggplot2")
```

The dosing compartment (`depot`, drawn with a heavy border) is on top and feeds
the central compartment.  The peripheral compartment, which exchanges mass with
`central` in both directions, is to the left, and elimination leaves the
central compartment downwards.

## Drawing engines

The same graph can be drawn several ways, chosen with `engine`:

- `"DiagrammeR"` draws an interactive Graphviz widget with the 'DiagrammeR'
  package.  This is the default when 'DiagrammeR' is installed.
- `"ggplot2"` returns a `ggplot` object that can be themed, saved with
  `ggplot2::ggsave()` or combined with other plots.  It needs no extra
  packages.
- `"dot"` returns the Graphviz DOT source as a string, to edit or to render
  with any Graphviz tool.

```{r, eval = hasDiagrammeR}
modelDiagram(two.cmt, engine = "DiagrammeR")
```

```{r}
cat(modelDiagram(two.cmt, engine = "dot"))
```

The default engine can be set once per session:

```{r, eval = FALSE}
options(nlmixr2plot.diagram.engine = "ggplot2")
```

Arrows can be labeled with the model terms that drive them:

```{r}
modelDiagram(two.cmt, engine = "ggplot2", labels = TRUE)
```

## The graph behind the diagram

`modelGraph()` returns the parsed graph, which is useful for checking how the
equations were interpreted (or for drawing it some other way):

```{r}
g <- modelGraph(two.cmt)
g
```

The `nodes` data frame gives each compartment's role and layout position, and
the `edges` data frame lists every flow.  A graph can be passed straight to
`modelDiagram()` or `plot()`:

```{r, eval = FALSE}
plot(g, engine = "ggplot2")
```

### How the equations are read

Each `d/dt()` equation is split into signed additive terms (products are
distributed over sums), and then:

- **Mass transfer**: a term subtracted from one compartment and added,
  identically, to another moves mass between them.  The order of factors does
  not matter (`ka*depot` matches `depot*ka`), and first-order, zero-order and
  enzyme-driven rates are all recognized.
- **Elimination**: a remaining loss that contains the compartment's own
  amount (like `cl/v*central` or Michaelis-Menten `vmax*C/(km + C)`).
- **Input**: a remaining gain that does not depend on any compartment (like a
  zero-order `kin`).
- **Interaction**: a term that depends on another compartment without moving
  mass, like an effect compartment or a pharmacodynamic stimulation or
  inhibition.  It is drawn dashed.  Its direction comes from the equation:
  inhibition when the term decreases as the driving compartment increases
  (including `1 - emax*C/(ec50 + C)`, `kin/(1 + C)` and `exp(-k*C)` forms),
  stimulation when it increases, and "modulation" when the direction cannot be
  determined.

Intermediate variables (like `cp <- central/v`) are followed, `if`/`else`
blocks and `ifelse()` keep their conditions, and residual error lines are
ignored.

## Pharmacodynamic models

Compartments that interact with the pharmacokinetic model without mass
transfer go to the right, with their own inputs above and outputs below.  Here
is an indirect response (turnover) model where the drug inhibits the
production of the response:

```{r}
pk.turnover <- function() {
  ini({
    tktr <- log(1)
    tka <- log(1)
    tcl <- log(0.1)
    tv <- log(10)
    poplogit <- 2
    tec50 <- log(0.5)
    tkout <- log(0.05)
    te0 <- log(100)
    prop.err <- 0.1
    pdadd.err <- 10
  })
  model({
    ktr <- exp(tktr)
    ka <- exp(tka)
    cl <- exp(tcl)
    v <- exp(tv)
    emax <- expit(poplogit)
    ec50 <- exp(tec50)
    kout <- exp(tkout)
    e0 <- exp(te0)
    DCP <- center / v
    PD <- 1 - emax * DCP / (ec50 + DCP)
    effect(0) <- e0
    kin <- e0 * kout
    d/dt(depot) <- -ktr * depot
    d/dt(gut) <- ktr * depot - ka * gut
    d/dt(center) <- ka * gut - cl / v * center
    d/dt(effect) <- kin * PD - kout * effect
    cp <- center / v
    cp ~ prop(prop.err)
    effect ~ add(pdadd.err)
  })
}
modelDiagram(pk.turnover, engine = "ggplot2")
```

The transit compartment `gut` stacks between the dosing compartment and the
central compartment.  The response has a zero-order input (`kin`) above it, an
elimination below it, and a dotted arrow from `center`: the drug *inhibits*
the response, because `PD` decreases as the concentration increases.  With
'DiagrammeR' the inhibition is drawn with a "tee" arrow head:

```{r, eval = hasDiagrammeR}
modelDiagram(pk.turnover, engine = "DiagrammeR")
```

A more involved example, with a metabolite, two peripheral compartments, an
effect compartment and a response driven by the effect compartment:

```{r}
pkpd <- rxode2::rxode2({
  C2 <- centr / V2
  C3 <- peri / V3
  C4 <- peri2 / V4
  d/dt(depot) <- -KA * depot
  d/dt(centr) <- KA * depot - CL * C2 - Q * C2 + Q * C3 - Q2 * C2 + Q2 * C4 -
    kmet * centr
  d/dt(peri) <- Q * C2 - Q * C3
  d/dt(peri2) <- Q2 * C2 - Q2 * C4
  d/dt(met) <- kmet * centr - kelm * met
  d/dt(ce) <- ke0 * (C2 - ce)
  d/dt(resp) <- kin - kout * (1 - ce / (ec50 + ce)) * resp
})
modelDiagram(pkpd, engine = "ggplot2")
```

## Target-mediated drug disposition

Binding and unbinding are recognized as mass transfer from both binding
partners into the complex (and back):

```{r}
tmdd <- rxode2::rxode2({
  d/dt(central) <- -kel * central - kon * central * target + koff * complex
  d/dt(target) <- ksyn - kdeg * target - kon * central * target +
    koff * complex
  d/dt(complex) <- kon * central * target - koff * complex - kint * complex
})
modelGraph(tmdd, dosing = "central")
modelDiagram(tmdd, dosing = "central", engine = "ggplot2")
```

## Dosing compartments

Dosing compartments are detected from the dosing records of the data: for a
fitted model this is the data it was fit to, and for other models it can be
supplied with `data`.  Both numeric and named `cmt` values are understood:

```{r}
d <- data.frame(
  id = 1, time = c(0, 0, 1, 2),
  amt = c(100, 50, 0, 0), evid = c(1, 1, 0, 0),
  cmt = c("depot", "central", "central", "central"), dv = 0
)
modelGraph(two.cmt, data = d)$nodes
```

With no data, the first compartment (`rxode2`'s default dosing compartment) is
assumed.  The dosing compartments can also be given directly:

```{r}
modelDiagram(two.cmt, dosing = "central", engine = "ggplot2")
```

For a fit, `modelDiagram(fit)` uses the fitted model and data:

```{r, eval = FALSE}
fit <- nlmixr2(two.cmt, nlmixr2data::theo_sd, est = "focei")
modelDiagram(fit)
```

## `linCmt()` models

Models written with `linCmt()` are converted to their ODE form with
`rxode2::linToOde()` before they are diagrammed (this requires a version of
'rxode2' that provides it):

```{r, eval = "linToOde" %in% getNamespaceExports("rxode2")}
one.cmt <- function() {
  ini({
    tka <- 0.45
    tcl <- 1
    tv <- 3.45
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka)
    cl <- exp(tcl)
    v <- exp(tv)
    linCmt() ~ add(add.sd)
  })
}
modelDiagram(one.cmt, engine = "ggplot2")
```

## Limitations

The diagram is only as good as the parsing of the equations, so a few
conventions matter:

- Mass transfer is only detected when the same term is subtracted from the
  source and added to the destination.  A transfer scaled in only one of the
  equations (for example a volume or stoichiometric conversion) is drawn as an
  elimination plus an interaction.
- A gain driven only by another compartment (like `ke0*cp` into an effect
  compartment) is represented by the dashed interaction arrow alone.
- The layout is automatic; for publication-quality tweaks, start from the DOT
  source (`engine = "dot"`) or the `ggplot` object and adjust it.
