---
title: "ODE block diagrams"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ODE block diagrams}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
hasDiagrammeR <- requireNamespace("DiagrammeR", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(nlmixr2plot)
library(rxode2)
```

`nlmixr2plot` can draw the ordinary differential equations (ODEs) of an
rxode2/nlmixr2 model as a control-systems block diagram. Call `plot()` on
the model's user interface object, e.g. `plot(rxode2(model))` or
`plot(model())`.

The diagram follows the classic recipe for turning an ODE into a
simulation scheme:

1. **Isolate the highest derivative.** rxode2 ODEs are already first
   order (`d/dt(x) <- ...`), so each state gets exactly one integrator
   ($1/s$).
2. **Integrators.** The integrator's input is `d/dt(x)` and its output is
   the state `x`.
3. **Summing junction.** A $\Sigma$ node in front of each integrator
   combines the additive terms of `d/dt(x)`. Each input is marked `+` or
   `-`.
4. **Gains and feedback.** Terms that are linear in one state are drawn as
   gain blocks (triangles) tapped from that state's integrator output. When
   a state feeds its own derivative (or an earlier state's), the edge is
   drawn as a dashed feedback edge.

Terms that don't depend on any state become external inputs (dashed
boxes). All other terms become nonlinear function blocks (purple),
wired from every state they depend on. This includes dependencies through
intermediate variables and `if()` conditions.

The diagram also shows rxode2's dosing and state features:

| Model code             | Diagram                                              |
|------------------------|------------------------------------------------------|
| dosing                 | green `Dose` input into the compartment's $\Sigma$   |
| `alag(cmt)`/`lag(cmt)` | orange delay block $e^{-s \cdot t_{lag}}$            |
| `f(cmt)`/`F(cmt)`      | green bioavailability gain `F`                       |
| `rate(cmt)`            | blue zero-order infusion block (modeled rate)        |
| `dur(cmt)`             | blue zero-order infusion block (modeled duration)    |
| `cmt(0)`               | initial condition shown on the integrator            |
| endpoints (`~`)        | grey output nodes, with the residual error model     |

`plot()` needs the 'DiagrammeR' package. Without it, `odeDiagram()` still
builds the diagram as data, which you can print or export as Graphviz DOT
(see the end of this vignette).

## One-compartment model with first-order absorption

Here the depot has a bioavailability and a lag time. The central
compartment has an initial amount and a modeled infusion duration.

```{r onecmt}
one.compartment <- function() {
  ini({
    tka <- 0.45
    tcl <- 1
    tv <- 3.45
    eta.ka ~ 0.6
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl)
    v <- exp(tv)
    d/dt(depot) <- -ka * depot
    d/dt(center) <- ka * depot - cl / v * center
    f(depot) <- 0.8
    alag(depot) <- 0.5
    center(0) <- 1
    dur(center) <- 2
    cp <- center / v
    cp ~ add(add.sd)
  })
}
```

```{r onecmt-plot, eval = hasDiagrammeR}
plot(one.compartment())
```

Reading the diagram:

- `ka * depot` shows up twice. Its gain feeds back negatively into the
  depot's $\Sigma$ and positively into the central compartment's $\Sigma$.
- `cl/v` is negative feedback on the central compartment.
- The depot dose passes through the lag time and then the bioavailability.
- The central compartment dose passes through the modeled infusion duration.

## Two-compartment `linCmt()` models

`linCmt()` models are first converted to their ODE form with
`rxode2::linToOde()`, so they are drawn the same way.

```{r twocmt}
two.compartment <- function() {
  ini({
    lka <- 0.45
    lcl <- 1
    lv <- 3.45
    lq <- 0.5
    lvp <- 4
    add.sd <- 0.7
  })
  model({
    ka <- exp(lka)
    cl <- exp(lcl)
    v <- exp(lv)
    q <- exp(lq)
    vp <- exp(lvp)
    rate(central) <- 5
    linCmt() ~ add(add.sd)
  })
}
```

```{r twocmt-plot, eval = hasDiagrammeR}
plot(two.compartment())
```

## Nonlinear elimination with an indirect-response effect

Michaelis–Menten elimination is not linear in the state, so it is drawn as
a function block. The turnover model's `kin` does not depend on any state,
so it is an external input. The `kout * (1 + cp) * eff` term depends on
both `eff` and, through `cp`, on `center`. The block lists the
intermediate's definition.

```{r pkpd}
pkpd <- function() {
  ini({
    vmax <- 10
    km <- 2
    v <- 10
    kin <- 1
    kout <- 0.1
    add.sd <- 0.1
    prop.sd <- 0.1
  })
  model({
    cp <- center / v
    d/dt(center) <- -vmax * cp / (km + cp)
    d/dt(eff) <- kin - kout * (1 + cp) * eff
    eff(0) <- kin / kout
    cp ~ prop(prop.sd)
    eff ~ add(add.sd)
  })
}
```

```{r pkpd-plot, eval = hasDiagrammeR}
plot(pkpd())
```

## Transit compartments

A chain of transit compartments shows as a series of integrators, each
with its own feedback loop and a forward gain to the next compartment.

```{r transit}
transit <- function() {
  ini({
    ktr <- 1
    ka <- 1
    cl <- 1
    v <- 10
    add.sd <- 0.1
  })
  model({
    d/dt(depot) <- -ktr * depot
    d/dt(transit1) <- ktr * depot - ktr * transit1
    d/dt(transit2) <- ktr * transit1 - ktr * transit2
    d/dt(gut) <- ktr * transit2 - ka * gut
    d/dt(center) <- ka * gut - cl / v * center
    cp <- center / v
    cp ~ add(add.sd)
  })
}
```

```{r transit-plot, eval = hasDiagrammeR}
plot(transit())
```

## Controlling the diagram

By default, doses are drawn for the first state and for every state with
a bioavailability, lag time, rate or duration. Use `doses=` to choose the
dosed compartments; `character(0)` hides dosing entirely.
`showZeroIni = TRUE` labels the integrators that have no explicit initial
condition with `x(0) = 0`.

```{r options, eval = hasDiagrammeR}
plot(transit(), doses = c("depot", "center"), showZeroIni = TRUE)
```

## Working with the diagram as data

`odeDiagram()` returns the diagram as `nodes` and `edges` data frames. It
accepts a model function, a `rxUi` object or a fitted nlmixr2 model.
Printing it gives a text summary of each state's inputs.

```{r data}
d <- odeDiagram(one.compartment)
d
head(d$nodes)
head(d$edges)
```

`as.character()` gives the Graphviz DOT source. You can render it with
any Graphviz tool, or pass it to `DiagrammeR::grViz()` yourself.

```{r dot}
cat(substr(as.character(d), 1, 600), "...\n")
```
